---
author: 'Keanu Rochette '
date: '`r format(Sys.Date())`'
title: 'HW 11 - GAMs '
output: 
  html_document: 
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE)
```

## Load libraries
```{r}
# Data manipulation and visualization 
library(tidyverse)
library(here)
library(janitor)

# Data analysis/statistics 
library(car) #ANOVA
library(ggResidpanel) #residual panel
library(ggeffects)
library(effects) #dependency for ggeffects
library(MuMIn)
library(gam)
library(mgcv)

```


## Load data

```{r}
FCM <- read_csv(here("HW11", "data", "HOT_cyto_counts.csv"))
MLDS <- read_csv(here("HW11", "data", "hot_mlds.csv"))

FCM <- clean_names(FCM)
MLDS <- clean_names(MLDS)

```

```{r}
# reformating the date and extracting the day of the year
# pressure turned negative for easier visualization later on 
FCM <- FCM %>% mutate(cal_date = mdy(date)) %>% relocate(cal_date, .after= date) %>% 
  mutate(doyear = yday(cal_date)) %>% relocate(doyear, .after= cal_date) %>% 
  mutate(press = -1*press)
```


## Question 1 

```{r, fig.width= 10, fig.height=6}
# making models of the microbes as a fucntion of pressure only 
proxdepth <- gam(sqrt(pro) ~ s(press), data= FCM)
hbactxdepth <- gam(hbact ~ s(press), data= FCM)
picoeukxdepth <- gam(sqrt(picoeuk)~ s(press), data= FCM)

par(mfrow = c(1,3))
plot(proxdepth, main = "pro", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "green")
plot(hbactxdepth, main = "hbact", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "red")
plot(picoeukxdepth, main= "pico euk", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "royalblue")
```




```{r, fig.width= 10, fig.height=6}
# making models of the microbes as a fucntion of day of year only 
proxday <- gam(pro ~ s(doyear), data= FCM)
hbactxday <- gam(hbact ~ s(doyear), data= FCM)
picoeukxday <- gam(sqrt(picoeuk)~ s(doyear), data= FCM)

par(mfrow = c(1,3))
plot(proxday, main = "pro", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "green")
plot(hbactxday, main = "hbact", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "red")
plot(picoeukxday, main = "pico euk", residuals = TRUE, cex = 4, lwd = 3, shade = TRUE, shade.col = "royalblue")
```

### Comprehesion models for each microbe

```{r}
# making the comprehensive model with pressure and day of year for prochlorococcus
pro2D <- gam(sqrt(pro) ~ s(doyear, press), data= FCM) 
pro2D_X <- gam(sqrt(pro) ~ s(doyear) + s(press), data= FCM) 
summary(pro2D)
``` 

```{r}
anova(pro2D, pro2D_X)
```

#### Interpretation

The F value of the anova is significant which means that the interaction between depth and day of year explains the variation in abundance more than the variables alone.


```{r}
# making the comprehensive model with pressure and day of year for het bacteria
hbact2D <- gam(hbact ~ s(doyear, press), data= FCM) 
hbact2D_X <- gam(hbact ~ s(doyear) + s(press), data= FCM) 
summary(hbact2D)
``` 

```{r}
anova(hbact2D, hbact2D_X)
```
#### Interpretation

The F value of the anova is significant which means that the interaction between depth and day of year explains the variation in abundance more than the variables alone.


```{r}
# making the comprehensive model with pressure and day of year for pico eukaryotes
picoeuk2D <- gam(sqrt(picoeuk) ~ s(doyear, press), data= FCM) 
picoeuk2D_X <- gam(sqrt(picoeuk) ~ s(doyear) + s(press), data= FCM) 
summary(picoeuk2D)
``` 

```{r}
anova(picoeuk2D,picoeuk2D_X)
```
#### Interpretation

The F value of the anova is significant which means that the interaction between depth and day of year explains the variation in abundance more than the variables alone.


```{r, fig.width=10, fig.height=6}
par(mfrow = c(1,3))
plot(pro2D, main = "pro", residuals = TRUE, cex = 4, lwd = 3, cex.main = 3)
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
plot(hbact2D, main = "hbact", residuals = TRUE, cex = 4, lwd = 3, cex.main = 3)
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
plot(picoeuk2D, main = "pico euk", residuals = TRUE, cex = 4, lwd = 3, cex.main = 3)
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
```

```{r, fig.width=12, fig.height=6}
par(mfrow = c(1,3))
vis.gam(pro2D, view = c("doyear", "press"), plot.type = "contour")
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
vis.gam(hbact2D, view = c("doyear", "press"), plot.type = "contour")
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
vis.gam(picoeuk2D, view = c("doyear", "press"), plot.type = "contour")
abline(v = 60, col = "pink",lwd = 3) #spring
abline(v = 152, col = "green",lwd = 3) #summer
abline(v = 244, col = "orange",lwd = 3) # fall
abline(v = 335, col = "skyblue",lwd = 3) # winter
```


## Question 2

```{r}
FCM_long <- FCM %>% 
  pivot_longer(cols = hbact:picoeuk,
                     names_to = "microbe", 
                     values_to = "micr_conc") %>% 
  filter(microbe != "syn") %>% 
  mutate(microbe = factor(microbe, levels= c("pro", "hbact", "picoeuk"))) 

```

### Setting up the model
```{r}
microbe_model <- gam(micr_conc ~ s(doyear, press, by = microbe) + microbe, data= FCM_long) 
summary(microbe_model)
```

```{r, fig.width=12, fig.height=6}
# I don't know why my picoeukaryotes plot shows a linear trend instead of a contour plot...
par(mfrow = c(1,3))
plot(microbe_model, residuals = TRUE, cex = 4, lwd = 3, cex.main = 3)
```



### Interpretation
**(a) Do the different kinds of microbes have different mean abundances?**  
Yes, we saw that the estimated intercepts are different with the summary(microbe_model). 

**(b) Do the different kinds of microbes have different average depth distributions (i.e., averaging over time)?**

```{r}
# we can compare the full model to a model in which only the depth (pressure) is accounted for in the variation of abundance of the microbes
microbe_depth <- gam(micr_conc ~ s(press, by = microbe) + microbe, data= FCM_long) 
```

```{r}
par(mfrow = c(1,3))
plot(microbe_depth)
```

```{r}
# The F value of the anova is significant for pro and hbact which means that they are differently distributed on average by depth when averaging over time.
summary(microbe_depth)
```


**(c) Do the different kinds of microbes have different average seasonal dynamics (i.e., averaging over depths)**

```{r}
# we can make a model that only accounts for the day of the year as a predictor for the variation in microbe abundance
microbe_doyear <- gam(micr_conc ~ s(doyear, by = microbe) + microbe, data= FCM_long) 
```

```{r}
par(mfrow = c(1,3))
plot(microbe_doyear)
```

```{r}
# the F value of the anova is only significant for hbact, which means that hetero trophic bacteria are distributed differently throughout the year. 
summary(microbe_doyear)
```


## Question 3 

```{r}
MLDS<- MLDS %>% rename(cruise = crn)
```

```{r}
full_df <- left_join(FCM, MLDS, by = c("cruise", "date"))
```


```{r}
HOT <- full_df %>% select(-julian, -syn, -botid, -x1) %>% 
  # we only want the data of the top 45m 
  filter(press <=45) %>%  
  filter(mean  <=45) %>% 
  pivot_longer(cols = hbact:picoeuk,
                     names_to = "microbe", 
                     values_to = "micr_conc") %>% 
  mutate(microbe = factor(microbe, levels= c("pro", "hbact", "picoeuk"))) %>% 
  rename(mixlay = mean)
  
```

### Mixed Layer
```{r}
#ML_gam <- gam(micr_conc ~ s(press, by = microbe)+ s(doyear, mixlay, by = microbe) + microbe, data = HOT)
ML_gam <- gam(micr_conc ~ s(mixlay, by = microbe) + microbe, data = HOT)
```

```{r}
summary(ML_gam)
```


```{r, fig.width=10, fig.height=5}
par(mfrow = c(1,3))
plot(ML_gam)
```

#### Interpretation
We can see that the heterotrophic bacteria are significantly associated with the variation in mixed layer depth. There is a peak of hbact abundance around 32m and a low in hbact abundance at 39m. 

### Chlorophyll
```{r}
chl_gam <- gam(micr_conc ~ s(chl, by = microbe) + microbe, data = HOT)
```

```{r}
summary(chl_gam)
```


```{r, fig.width=10, fig.height=5}
par(mfrow = c(1,3))
plot(chl_gam)
```

#### Interpretation

- Prochlorococcus abundance vary with chlorophyll concentration. It looks like the maximum Prochlorococcus abnundance can be found wit 0.12ug/L of chlorophyll .
- Heterotrophic bacteria varies similarly with variation of chlorophyll with a peak of hbact abundance at 0.10ug/L of chlorophyll. 











