---
author: 'Keanu Rochette '
date: '`r format(Sys.Date())`'
title: HW3 - Coding practice
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE)
```

## Load Libraries
```{r}
# Data manipulation and visualization 
library(tidyverse)
library(here)
library(janitor)
library(ghibli)
library(RColorBrewer)
library(gt)
library(gtExtras)

# Data analysis/statistics 
library(car) #ANOVA
library(ggeffects)
library(effects) #dependency for ggeffects
library(ggResidpanel) #residual panel
library(GGally)

```


## Load Data 
```{r}
fish <- read_csv(here("HW3","data","CRCP_reef_fish.csv"))
```

```{r}
fish <- fish %>% clean_names()
```

```{r}
# I will be using the same theme elements for all my plots to I can automate it using this 
nunu_theme <- theme_bw() + 
  theme(plot.title = element_text(size=14, face = "bold"), 
        plot.subtitle = element_text(size=12),
        axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12),
        strip.text.x = element_text(size = 12, face = "bold"),
        legend.title=element_text(size=12, face = "bold"),
        legend.text=element_text(size=12),
        #axis.text.x = element_text(angle= 45, hjust= 1),
        panel.background = element_rect(fill = "azure1"))
```

## Question 1: summary statistics 

_Identify the top twelve species in terms of these three statistics: (1) mean abundance, (2) maximum abundance, (3) standard deviation of abundance. In other words, which species are most abundant on average, which reach the highest abundances, and which have the most variable abundances?_

```{r}
# create a summary table with average abundance, the max count and the standard deviation
fish_summary_stats <- fish %>% group_by(taxonname) %>% 
  summarize(avg = mean(count, na.rm = T),
            max_abun = max(count, na.rm = T),
            std_dev = sd(count, na.rm = T))
```

### A- Top 12 fish species in mean abundance

```{r}
# arrange the average abundance in decreasing order and keep the top 12
fish_summary_stats %>% arrange(desc(avg)) %>% 
  head(12) %>% gt() %>% 
  gt_highlight_cols(avg,  fill= "orange", alpha = 0.4)
```

### B- Top 12 fish species in max abundance
```{r}
# arrange the max abundances in decreasing order and keep the top 12
fish_summary_stats %>% arrange(desc(max_abun)) %>% 
  head(12) %>% gt() %>% 
  gt_highlight_cols(max_abun,  fill= "orange", alpha = 0.4)
```
### C- Top 12 fish species in highest standard deviation
```{r}
# arrange the standard deviations in decreasing order and keep the top 12
fish_summary_stats %>% arrange(desc(std_dev)) %>% 
  head(12) %>% gt() %>% 
  gt_highlight_cols(std_dev,  fill= "orange", alpha = 0.4)
```

## Question 2: Depth 
__Using the top twelve species based on mean abundance, plot a scatterplot of count vs. depth (the column is named ‘depth’). Add a smoother to help visualize mean count vs. depth, put all twelve species on a single plot (with twelve panels), and make sure each panel is titled using the species name. Furthermore, do not arrange the panels in alphabetical order — instead, arrange them in order of mean abundance, so that the most abundant species is in the top left panel, and remaining panels are in order of mean abundance. Do it in a for loop and in ggplot.__

```{r}
# arrange the average abundance in decreasing order and keep the top 12 name of fish species
top_avg_abun <- fish_summary_stats %>% arrange(desc(avg)) %>% 
  head(12) %>%  pull(taxonname)
```

### Graph with a for loop 

```{r, warning= FALSE, fig.width= 12, fig.height=6}
#for loop to graph the correlation between fish count and depth

par(mfrow= c(3,4)) #creates a panel of 3cols x 4rows figure
for(i in 1:length(top_avg_abun)){
  
  data <- fish[fish$taxonname == top_avg_abun[i],] #filters the fish species based on their order form highest to lowest average abundance
  
  plot(sqrt(count) ~ depth, data = data, # sqrt transform the count and compare to depth
       col = "#616161", # change color of the points
       main = top_avg_abun[i]) # specify the title of the plot to be fish species name
  
   with(data, lines(loess.smooth(depth,sqrt(count)), col = "blue", lwd = 3))
   # creates the trend line with specific color and linewidth
  
}

```


### Graphing with ggplot 
```{r, fig.width= 12, fig.height=6}
#plot the correlation between fish count and depth with ggplot 

fish %>% filter(taxonname %in% top_avg_abun) %>% 
  mutate(taxonname = fct_reorder(taxonname, 
                                 .x= count, .fun = mean,
                                 .desc = TRUE)) %>% 
  ggplot(aes(x= depth, y = count)) +
  geom_point(alpha = 0.7)  +
  geom_smooth(linewidth = 2)+
  facet_wrap(.~taxonname, scales = "free")+
 scale_y_sqrt() +
  labs(x= "Depth (m)",
       y = "Count (sqrt)")+
  nunu_theme
  
```

```{r}
# name of the top 12 fish species
top_avg_abun
```


### Answer

Fish species seem to have particular depth preference:

- **Surface**: Acanthurus nigrofuscus, Zebrasoma flavescens, Acanthurus leucopareius, Thalassoma duperrey.
- **Mid-depth**: Chromis vanderbilti, Chromis agilis, Ctenochaetus strigosus, Paracirrhites arcatus, Acanthurus leucopareius, Decapterus macarellus.
- **Deep**: Lutjanus kasmira, Pristiapogon kallopterus, Chromis hanui.

## Question 3

```{r}
# create a vector that keep only the name of the top 5 most abundant fish species on average 
top_abun_fish <- fish_summary_stats %>% arrange(desc(avg)) %>% 
  head(5) %>%  pull(taxonname)
```

### A- Plotting the most abundant species (on average)

_Make a new plot that shows abundance vs. depth for the top 5 species, including smoothers, but this time put all of the species on the same scatterplot and distinguish them with different colors. This time you don’t need to use two different approaches to make the plot — one approach will suffice. What is your interpretation of this plot?_

```{r fig.width= 12, fig.height=6}

fish %>% filter(taxonname %in% top_abun_fish) %>% #keeping only the top 12 most abundant fish 
  mutate(taxonname = fct_reorder(taxonname, 
                                 .x= count, .fun = mean,
                                 .desc = TRUE)) %>% #order the taxoname by abundance
  ggplot(aes(x= depth, y = count,
             color = taxonname, group = taxonname))+
  geom_point()+
  geom_smooth(method = "loess")+ # choose smoothing method
  scale_color_brewer(palette = "Set1")+ # choose colot palette
  scale_y_sqrt()+ # sqrt transform the y axis
  nunu_theme
```

#### Answer
Looking at the trend lines, it looks like fish species have different depth preference. It appears to be sequential as well.  

- **Chromis vanderbilti**'s abundance is more correlated to the surface (0-20m). 
- **Chromis agilis**'s abundance correlates with deeper depth (15-25m).
- **Ctenochaetus strigosus** also prefers mid to deeper depth, with a maximum at around 20m, but the overall trend it fairly flat.
- **Zebrasoma flavescens** & **Acanthurus nigrofuscus**'s abundances does not seem to correlate with a particular depth. 

It looks like the top most abundant fish species have a preference in depth of occurrence. Theure seem to be a sequential distribution of the fish species with depth, as if each species is more predominant in their "territory". However some exception exist such as Chromis ovalis, which seem to be occurring homogeneously across depth. 

### B- Linear model 

_Fit a linear model that tests whether these 5 species have different depth niches. Based on residual diagnostic plots of this model, and a plot of the fitted effects, do you think a linear model is a good approach for testing this question? If not, why not?_

```{r}
top_5_fish <- fish %>% filter(taxonname %in% top_abun_fish) #only keep the top 5 fish species
```


```{r}
#setting up a linear model. depth and taxonname are predictors and sqrt(count) is a response variable
top_fish_lm <- lm(sqrt(count) ~ depth*taxonname, data= top_5_fish)

summary(top_fish_lm)
```

```{r}
# Anova show significant differences between fish species and in the interaction between taxonname and depth 
Anova(top_fish_lm) 

```


```{r}
# plot the residuals 
resid_panel(top_fish_lm, plots = c("resid", "qq", "lev", "hist"))
```

```{r fig.width=12, fig.height=6}
# plot the effects plot
plot(allEffects(top_fish_lm, terms = c("depth", "taxonname")))
```

```{r}
# plot the effects plot

plot(ggeffect(top_fish_lm, terms = c("depth", "taxonname")))
```


#### Answer

A linear model might not be the best statistical tool to understand the distribution of fish species with depth.  
The **residual plot* show that the residual points are not evenly distributed on each side of the line which may indicate that the data point are not independent from one another.  
The **Q-Q plot** shows that the residual points do not follow a regression line.  
The **histogram** shows that the data is not distributed normally. 
