---
author: 'Keanu Rochette '
date: '`r format(Sys.Date())`'
title: HW5- Fish Survey Count
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE)
```

## Load Libraries
```{r}
# Data manipulation and visualization 
library(tidyverse)
library(here)
library(janitor)
library(patchwork)
#library(grid)
#library(gridExtra)
#library(ghibli)
#library(RColorBrewer)
#library(gt)
#library(gtExtras)

# Data analysis/statistics 
library(car) #ANOVA
library(ggeffects)
library(effects) #dependency for ggeffects
library(ggResidpanel) #residual panel
library(GGally)
library(arm) #discrete.histogram() function
library(MASS) #required for arm package
```

## Load the functions

```{r}
dispersion <- function (model){
  sum(residuals(model, type = "pearson")^2)/(length(model$y)-length(model$coefficients))
}
```



## Load data
The variables with are interested in are:
- turf algae (column ‘ta’)
- hard coral (column ‘hard_coral’) 
- sand (column ‘sand’)
- crustose coralline algae (column ‘cca’)
```{r}
fish <- read_csv(here("HW5","data","CRCP_Reef_Fish_Surveys_kole.csv"))
```

## Question 1: Predictor Correlations

```{r}
fish1 <- fish[,c("ta","hard_coral", "sand", "cca")]
ggpairs(fish1)

```


## Question 2: Poisson distribution with single predictor model
_For each of the four predictors, make single-predictor models where the response variable is kole counts. Consider which of the predictors could be transformed to reduce skew along the x-axis. For educational purposes, start by using a poisson distribution for the counts._

### Turf Algae

```{r}
#Based on the frequency distribution of the turf algae (ta) data, the predictor does not seem to be suited for the Poisson distribution. 
par(mfrow = c(1,2))
discrete.histogram(fish$count)
discrete.histogram(fish$ta)


```

```{r}
ta_lm <- lm(count ~ ta, data = fish)
summary(ta_lm)

resid_panel(ta_lm, plots = c("resid", "qq", "lev", "hist"))
```

```{r}
ta_glm <- glm(count ~ ta, family= poisson, data = fish)
summary(ta_glm)
resid_panel(ta_glm, plots = c("resid", "qq", "lev", "hist"))
resid_xpanel(ta_glm)

```


```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(ta_glm) %>% plot(log_y = TRUE)
ggeff



```

```{r}
Anova(ta_glm)
```

```{r}
discrete.histogram(fish$ta, prob.col = 'orange', xlim = c(0, 100), ylim = c(0,0.18))
lines(0:25, dpois(0:25, lambda = mean(fish$count)), col = 'blue',
lwd = 4, type = 'b')
legend(12, 0.15, lwd = 4, col = 'blue', lty = 1, legend = "Poisson
prediction")
```


**Overdispersion**: Poisson and Binomial distributions makes assumptions on the variability of the data which is used for the calculations of likelihood. Larger variability will result in smaller Confidence Intervals (CI) and p-values, which is inaccurate.   
With a dispersion of 30.10 (>>>1), the data is over dispersed and it will result in smaller CI and p-values than it should. 
```{r}
dispersion(ta_glm)
```



### Hard Coral

```{r}
#Based on the frequency distribution of the hard coral data, the predictor seem to be suited for the Poisson distribution. 
par(mfrow = c(1,2))
discrete.histogram(fish$count)
discrete.histogram(fish$hard_coral)


```

```{r}
coral_lm <- lm(count ~ hard_coral, data = fish)
summary(coral_lm)

resid_panel(coral_lm, plots = c("resid", "qq", "lev", "hist"))
```

```{r}
coral_glm <- glm(count ~ hard_coral, family= poisson, data = fish)
summary(coral_glm)
resid_panel(coral_glm, plots = c("resid", "qq", "lev", "hist"))
resid_xpanel(coral_glm)

```


```{r}
ggeff <- ggeffect(coral_glm) %>% plot(log_y = TRUE)
ggeff

```

```{r}
Anova(coral_glm)
```

```{r}
discrete.histogram(fish$hard_coral, prob.col = 'orange', xlim = c(0, 75))
lines(0:25, dpois(0:25, lambda = mean(fish$count)), col = 'blue',
lwd = 4, type = 'b')
legend(12, 0.20, lwd = 4, col = 'blue', lty = 1, legend = "Poisson
prediction")
```

```{r}
dispersion(coral_glm)
```
With a dispersion of 28.13 (>>>1), the data is over dispersed and it will result in smaller CI and p-values than it should. 



### Sand

```{r}
#Based on the frequency distribution of the hard coral data, the predictor seem to be suited for the Poisson distribution. 
par(mfrow = c(1,2))
discrete.histogram(fish$count)
discrete.histogram(fish$sand)


```



```{r}
sand_glm <- glm(count ~ sand, family= poisson, data = fish)
summary(sand_glm)
resid_panel(sand_glm, plots = c("resid", "qq", "lev", "hist"))
resid_xpanel(sand_glm)

```


```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(sand_glm) %>% plot(log_y = TRUE)
ggeff

```

```{r}
Anova(sand_glm)
```

```{r}
discrete.histogram(fish$sand, prob.col = 'orange', xlim = c(0, 75))
lines(0:25, dpois(0:25, lambda = mean(fish$count)), col = 'blue',
lwd = 4, type = 'b')
legend(12, 0.20, lwd = 4, col = 'blue', lty = 1, legend = "Poisson
prediction")
```

```{r}
dispersion(sand_glm)
```
With a dispersion of 30.98 (>>>1), the data is over dispersed and it will result in smaller CI and p-values than it should. 


### CCA

```{r}
#Based on the frequency distribution of the hard coral data, the predictor seem to be suited for the Poisson distribution. 
par(mfrow = c(1,2))
discrete.histogram(fish$count)
discrete.histogram(fish$cca)


```


```{r}
cca_glm <- glm(count ~ cca, family= poisson, data = fish)
summary(cca_glm)
resid_panel(cca_glm, plots = c("resid", "qq", "lev", "hist"))
resid_xpanel(cca_glm)

```


```{r}
ggeff <- ggeffect(cca_glm) %>% plot(log_y = TRUE)
ggeff
```

```{r}
Anova(cca_glm)
```

```{r}
discrete.histogram(fish$cca, prob.col = 'orange', xlim = c(0, 75))
lines(0:25, dpois(0:25, lambda = mean(fish$count)), col = 'blue',
lwd = 4, type = 'b')
legend(12, 0.20, lwd = 4, col = 'blue', lty = 1, legend = "Poisson
prediction")
```

```{r}
dispersion(cca_glm)
```
With a dispersion of 30.83 (>>>1), the data is over dispersed and it will result in smaller CI and p-values than it should. 


### Answer: Relationship between Kole and the predictors
- There is a positive relationship between Kole and hard corals, and CCA. An increase in those predictors will result in an significant increase in Kole count.
- There is a negative relationship between Kole and turf algae, and sand. An increase in those predictors will result in an significant decrease in Kole count

## Question 3: dealing with overdispersion in a model
To deal with **over-dispersion**, we can use a **negative binomial** distribution or use **random effects**. 

### Turf Algae

```{r}
#negative binomial distribution 
ta_nb <- glm.nb(count ~ ta, data = fish)
summary(ta_nb)


```


```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(ta_nb) %>% plot(log_y=TRUE)
ggeff

```

```{r}
Anova(ta_nb)
```

**Overdispersion**: 
With a dispersion of 0.98, the overdispersion is fixed. 
```{r}
dispersion(ta_nb)
```

### Hard Corals

```{r}
#negative binomial distribution 
hc_nb <- glm.nb(count ~ hard_coral, data = fish)
summary(hc_nb)


```

```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(hc_nb) %>% plot(log_y=TRUE)
ggeff

```

```{r}
Anova(hc_nb)
```

**Overdispersion**: 
With a dispersion of 1.08, the overdispersion is fixed. 
```{r}
dispersion(hc_nb)
```

### Sand
```{r}
#negative binomial distribution 
sand_nb <- glm.nb(count ~ sand, data = fish)
summary(sand_nb)


```


```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(sand_nb) %>% plot(log_y=TRUE)
ggeff

```

```{r}
Anova(sand_nb)
```

**Overdispersion**: 
With a dispersion of 0.95, the overdispersion is fixed. 
```{r}
dispersion(sand_nb)
```

### CCA 
```{r}
#negative binomial distribution 
cca_nb <- glm.nb(count ~ cca, data = fish)
summary(cca_nb)


```


```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(cca_nb) %>% plot(log_y=TRUE)
ggeff

```

```{r}
Anova(cca_nb)
```

**Overdispersion**: 
With a dispersion of 0.99, the overdispersion is fixed. 
```{r}
dispersion(ta_nb)
```

### Answer
The dispersion values became closer to one using the negative binomial. This is because in this model, the variance is proportional to the mean and data points become corrected by a factor of "Phi".

## Question 4: Including all substrate types in a single model 

For this model, I tried a regular glm() with Poisson distribution and the dispersion was about 27. I decided to do negative binomial model to minimize the dispersion. 
```{r}
full_glm <- glm.nb(count ~ ta + hard_coral + sand + cca, data = fish)
summary(full_glm)
```



```{r}
# lecture notes suggest using plot(log.y = TRUE) but I get an error. 
ggeff <- ggeffect(full_glm) %>% plot(log_y=TRUE)

ggeff

```


```{r}
Anova(full_glm)
```


**Overdispersion**: 
With a dispersion of 0.99, the overdispersion is fixed. 
```{r}
full_glm <- glm.nb(count ~ ta + hard_coral + sand + cca, data = fish)
dispersion(full_glm)
```
### Diagnostic plots

The resid_panel function did not work with the glm.nb(). 


### Answer

- **Turf Algae**: The effects shifted from a negative slope to a positive slope. The Chi Square value is 1000x greater in the new model.
- **Hard Corals**: The effect slope remains positive but the magnitude of the effect is much greater in the new model. The Chi Square value is 100,000x greater in the new model.
- **Sand**: The effect slope shifted from a negative slope to a positive slope and the magnitude is greater in the new model. The Chi Square value is 10,000x greater in the new model.
- **CCA**: The effect slope remains positive but the magnitude of the effect is much greater in the new model. The Chi Square value is 100x greater in the new model.
- **Explanation to changes**: The model is now accounting for the all 4 benthic cover types and is able to consider the effects of each of the variable on one another instead of consider an individual effect. 


### Closing thoughts
Overall, it looks like corals and CCA are more associated with higher counts of Kole. We can possibly perform a linear regression and calculate the R-squared and p-values to estimate the significance of the correlation. 



